\CRANpkg{lg}: An R package for Local Gaussian Approximations

### 2021-09-09 otneim.tex
Started on the bibfile

\citet{tjos:huft:2013} propose the \emph{local Gaussian correlation} (LGC) as a new measure of statistical dependence between two stochastic variables $X_1$ and $X_2$, which has the following important property yet unrivaled in the literature: It can separate between positive and negative nonlinear dependence while still reducing to the ordinary Pearson correlation coefficient if $X_1$ and $X_2$ are jointly normally distributed. The R-package \CRANpkg{localgauss} \citep{bere:klep:tjos:2014} provides two important functions in this context; one that calculates the sample LGC based on observed values of $(X_1,X_2)$, and one that uses the estimated LGC to perform a local test of independence between $X_1$ and $X_2$ as described in detail by \citet{bere:tjos:2014}.

@@

Consider a random vector $X$ having the unknown probability density function $f_X(x)$. It is a standard task to estimate $f_X$ based on a random sample $X_1,\ldots, X_n$, and the statistical literature provides an abundance of methods to accomplish this. One may, for example, make the assumption that the unknown density function has a particular parametric form, $f_X \in F_{\theta}$, where $F_{\theta} = \{f(x;\theta), \theta \in \Theta\}$ is a family of probability density functions indexed by some parameter $\theta$, and where $\Theta$ is the parameter space. Under this assumption we will typically produce an estimate of the parameter $\theta$, written $\widehat \theta$, using the maximum likelihood method. The estimated probability density function is then given as $\widehat f_X(x) = f(x; \widehat\theta)$.

A different approach is to estimate $f_X(\cdot)$ without any prior parametric assumptions. The classical method for non-parametric density estimation is the kernel estimator 
$$\widehat f_X(x) = \frac{1}{nb}\sum_{i=1}^n K\left(\frac{X_i - x}{b}\right),$$
where $K$ is a symmetric density function (the kernel) and $b$ is a tuning parameter (the bandwidth) that controls the smoothness of the estimate $\widehat f_X(\cdot)$. See \citet{silv:1986} for an introduction to this topic. There is also a massive statistical literature on density estimation containing extensions and improvements to the classical methods to be used in various practical situations.

\citet{hjor:jone:1996} provide one such idea. They consider a parametric family $F_{\theta}$, but instead of searching for a single parameter value $\theta_0$, for which $f_X(x) = f(x;\theta_0)$ (or approximately so), they rather assert that different members of $F_{\theta}$ may approximate $f_X$ locally in different parts of its domain. In other words, they seek to estimate a parameter \emph{function} $\theta_0(x)$ for which $f_X(x) = f(x;\theta_0(x))$ (or approximately so), and do this by maximizing a \emph{local likelihood function} in each point $x$:
\begin{align}
\widehat\theta(x) 
&= \arg\max_{\theta \in \Theta} L_n(\theta, x) \nonumber\\ 
&= \arg\max_{\theta \in \Theta} \frac{1}{nb}\sum_{i=1}^n K\left(\frac{X_i - x}{b}\right)\log f(X_i; \theta) - \int \frac{1}{b}K\left(\frac{y - x}{b}\right)f(y;\theta)\,\textrm{d}y,
\label{eq:loclik}
\end{align}
where, again, $K$ is a symmetric density function and $b$ is a bandwidth parameter that controls the smoothness of the estimate. The second term in the local likelihood function is a penalty that ensures that the estimated density $\widehat f_X(x) = f(x; \widehat\theta(x))$ converges correctly to the true density function $f_X(x)$ as the sample size $n$ increases to infinity, and the bandwidth $b$ decreases towards zero. See \citet{hjor:jone:1996} for a detailed discussion about this construction.

\citet{tjos:huft:2013} consider the bivariate case $X = (X_1, X_2)$ and take $F_{\theta}$ to be the family of bivariate normal distributions consisting of densities on the form

\begin{align}
f(x;\theta) 
&= \psi(x_1, x_2;\mu_1, \mu_2, \sigma_1, \sigma_2, \rho)  \nonumber\\ 
&= \frac{1}{2\pi \sigma_1 \sigma_2\sqrt{1-\rho^2}} \nonumber \\
& \qquad \times\exp \left\{-\frac{1}{2(1-\rho^2)}\left(\frac{(x_1-\mu_1)^2}{\sigma_1^2}-2\rho\frac{(x_1-\mu_1)(x_2-\mu_2)}{\sigma_1 \sigma_2}+\frac{(x_2-\mu_2)^2}{\sigma_2^2}\right)\right\},
\label{eq:gaussian}
\end{align}
where $\theta = (\mu_1,\mu_2,\sigma_1,\sigma_2,\rho)$ is the vector of parameters. Using a sample $\{X_{1i}, X_{2i}\}$, $i = 1, \ldots, n$, they estimate $\theta$ locally in the point $x$ by maximizing the local likelihood function \eqref{eq:loclik}, producing
$$\widehat\theta(x) = (\widehat\mu_1(x), \widehat\mu_2(x), \widehat\sigma_1(x), \widehat\sigma_2(x), \widehat\rho(x)),$$
and take special interest in the estimated correlation function $\widehat\rho(x)$ (i.e. the LGC) because it serves as an attractive local measure of statistical dependence between $X_1$ and $X_2$. They show that the LGC reveals many types of nonlinear statistical dependence that are not captured by the ordinary (global) Pearson correlation coefficient. Furthermore, the LGC distinguishes between positive and negative dependence and reduces to the Pearson $\rho$ if $X_1$ and $X_2$ are jointly normal. We refer to \cite{tjos:huft:2013} for a detailed treatment of the theoretical foundations of the LGC as well as several examples, and rather present two simple illustrations at this point in order to demonstrate the concept.

\begin{figure}[t]
    \centering
    \begin{subfigure}[t]{0.48\textwidth}
      \includegraphics[width=\textwidth]{gaussian-example}
      \caption{Estimated local Gaussian correlations between two  jointly normal variables having correlation equal to 0.5, based on 1000 observations.}
    \end{subfigure}
    \begin{subfigure}[t]{0.48\textwidth}
      \includegraphics[width=\textwidth]{returns-example}
      \caption{Local Gaussian correlation between daily returns from the CAC40 and FTSE100 stock indices, based on 1000 consecutive trading days.}
    \end{subfigure}
    \caption{Two dependence maps}
    \label{fig:initial-example}
  \end{figure}

@@

Taking the LGC as a measure of dependence opens up for a number of possibilities to construct statistical tests. \citet{bere:tjos:2014} show that $\rho(x) \equiv 0$ implies that $X_1$ and $X_2$ are independent. They show further that independence between $X_1$ and $X_2$ implies $\rho(x) \equiv 0$ if the population values of the local mean and standard deviation functions satisfy the following conditions: $\mu_i(x_1, x_2) = \mu_i(x_i)$ and $\sigma_i(x_1, x_2) = \sigma_i(x_i)$ for $i=1,2$. Equivalence between independence and $\rho(x) \equiv 0$ holds in general if the observations have been suitably transformed according to a procedure presented later in this section. It follows then that departures from $\widehat\rho(x) \equiv 0$ may be taken as evidence against the hypothesis that $X_1$ and $X_2$ are statistically independent. \citet{bere:tjos:2014} formalize this notion by testing whether $\rho(x) \equiv 0$ for all $x\in S \subset \mathbb{R}^2$ using the test statistic

\begin{equation}
T_{n,b} = \int_S h\left(\widehat \rho(x) \right) \,\textrm{d}F_n(x)
\label{eq:test-statistic1}
\end{equation}
for some non-negative function $h$, for example $h(x) = x^2$ or $h(x) = |x|$. Critical values may be obtained by permutations of the data under the null hypothesis, and we demonstrate the implementation of this test using the \CRANpkg{lg}-package in Section \ref{chap:indtests}.

Consider next the stationary time series $\{X_t\}$. The autocorrelation function (ACF) $\rho_k = \rho(X_t, X_{t-k})$ is a well known concept for describing the serial dependence in the time series, but the ACF is, again, only capable to completely capture \emph{linear} serial dependence. \citet{laca:tjos:2017} seek to remedy this by rather calculating the the local correlation between $X_t$ and $X_{t-k}$. This leads to a test for serial independence in the natural way. In fact, this work is mainly a theoretical exercise in order to accommodate time series dependence. Testing for independence between $X_t$ and $X_{t-k}$ using observations $\{X_t, X_{t-k}\}_{t = k+1}^T$ leads to the same test statistic \eqref{eq:test-statistic1} and bootstrap procedure as the test for independence between $X_1$ and $X_2$ that we described above.

\citet{laca:tjos:2018} extend this problem to test for serial cross-dependence between two time series $\{X_t, Y_t\}$ by measuring the LGC between $X_t$ and $Y_{t-k}$. Departures from $\widehat\rho(x,y) \equiv 0$ is again taken as evidence against independence, and the test statistic (\ref{eq:test-statistic1}) provides an aggregate measure of this discrepancy in the specified region $S$. In this case, however, we can not obtain replicates of the test statistic under the null hypothesis by simple permutations of the data. \citet{laca:tjos:2018} suggest rather two block bootstrap procedures to this end, using fixed and random block sizes respectively. The tests for serial dependence, and serial cross-dependence are both implemented in the \CRANpkg{lg}-package, as we demonstrate in Section \ref{chap:indtests}.

We find another application of the local Gaussian approximation in the work by \citet{stov:tjos:huft:2014}, who measure and test for financial contagion. They define contagion by "a significant increase in cross-market linkages after a shock to one country" \citep[p. 2223]{forb:rigo:2002} and employ the LGC to quantify this potential linkage. The authors estimate the LGC on a grid $\{x_1, x_2\}_k$, $k=1,\ldots,K$ along the diagonal $D=\{(x_1,x_2): x_1 = x_2\}$ \emph{before} and \emph{after} some critical event in the financial markets, denoted as the crisis (C) and the non-crisis (NC) periods respectively. They compare the two estimates using the following test statistic;

$$T_{n,b}^D = \sum_{k=1}^K \left\{\widehat\rho_C(x_k, x_k) - \widehat\rho_{NC}(x_k, x_k) \right\}w(x_k,x_k),$$

where $w(\cdot, \cdot)$ is a weight function that serve the same purpose as the integration area $S$ in (\ref{eq:test-statistic1}). In this case, \citet{stov:tjos:huft:2014} show that a standard bootstrap will suffice in order to produce approximate replicates of $T_{n,b}^D$ under the null hypothesis of no financial contagion, and we demonstrate the implementation of this test using the \CRANpkg{lg}-package in Section \ref{chap:cont}.

Although the original work by \citet{hjor:jone:1996} provide a general framework for local likelihood density estimation using any $p$-variate parametric family as the local family, it is evident that the method may struggle in multivariate applications much in the same way as the kernel density estimator does. This is a consequence of the curse of dimensionality, the effect of which is sought remedied by an algorithm provided by \citet{otne:tjos:2017}. The idea is to fit the $p$-variate normal distribution $\psi(\mu, \Sigma)$ locally, where $\mu$ is the vector of $p$ expectations, and $\Sigma$ is the $p\times p$ covariance matrix (to which the correlation matrix $R$ corresponds), but under the following structural simplifications:

\begin{align}
\mu_i(x) &= \mu_i(x_1, \ldots, x_p) \stackrel{\textrm{def}}{=} \mu_i(x_i) \label{eq:simp-mu}\\
\sigma_i(x) &= \sigma_i(x_1, \ldots, x_p)  \stackrel{\textrm{def}}{=} \sigma_i(x_i) \label{eq:simp-sig} \\
\rho_{ij}(x) &= \rho_{ij}(x_1, \ldots, x_p) \stackrel{\textrm{def}}{=} \rho_{ij}(x_i, x_j). \label{eq:simp-rho}
\end{align}
\citet{otne:tjos:2017} estimate the local parameters above by first obtaining univariate marginal locally Gaussian fits (eqs. \ref{eq:simp-mu} and \ref{eq:simp-sig}), and then pairwise bivariate locally Gaussian fits (eq. \ref{eq:simp-rho}). In the second step the estimates $\widehat\mu_i(x_i)$, $\widehat\mu_j(x_j)$, $\widehat\sigma_i(x_i)$ and $\widehat\sigma_j(x_j)$ are kept fixed in the estimation of the pairwise local correlation. They argue further that the following transformation technique will produce better density estimates in many situations. The motivation for introducing the simplifications defined in equations \ref{eq:simp-mu}-\ref{eq:simp-rho} can be compared to the practical advantages of estimating additive regression models, where $\textrm{E}(Y) = f(x_1, \ldots, x_p) \stackrel{\textrm{def}}{=} f_1(x_1) + \cdots + f_p(x_p)$.

Denote by $F_i(x_i)$, $i = 1,\ldots, p$ the marginal distribution functions of the stochastic vector $X$, and by $\widehat F_i(x_i) = n^{-1}\sum_{i=1}^n 1(X_i \leq x_i)$ their empirical counterparts. They then estimate the density $f_Z(z)$ of the vector $Z = \{\Phi^{-1}(F_i(X_i))\}_{i=1,\ldots,p}$, in practice approximated by

\begin{equation}
\widehat Z = \{\Phi^{-1}(\widehat F_i(X_i))\}_{i=1,\ldots,p}, 
\label{eq:trans}
\end{equation}
and where $\Psi(\cdot)$ is the univariate standard normal cdf. In that case, they simplify the estimation problem even further and fix

\begin{equation}
\mu_i(z_i) \stackrel{\textrm{def}}{=} 0 \,\, \textrm{ and } \sigma_i(z_i) \stackrel{\textrm{def}}{=} 1, \,\, i = 1,\ldots,p,
\label{eq:simp}
\end{equation}
so that the only parameter functions left to estimate are the pairwise local Gaussian correlations $R(z) = \{\rho_{ij}(z_i, z_j)\}_{i<j}$. We use the notation $Z$, $z_i$ and $z_j$ to signify that the estimation is performed on the (approximate) standard normal scale, or $z$-scale for short. We can then estimate the joint density $f_Z(z)$ of $Z$ as
\begin{equation}
\widehat f_Z(z) = \psi(z; \mu(z) = 0, \sigma(z) = 1, R = \widehat R(z)),
\label{eq:transformed-density}
\end{equation}
where $\mu(z) = \{\mu_i(z)\}$ and $\sigma(z) = \{\sigma_i(z)\}$ for $i=1,\ldots,p$, and then substitute $f_Z$ for $\widehat f_Z$ in the following relation obtained by \citet{otne:tjos:2017} in order to estimate the unknown density $f_X$:
\begin{equation}
f(x) = f_{Z}\big(\Phi^{-1}(F_1(x_1)), \ldots, \Phi^{-1}(F_p(x_p))\big) \times \prod_{i=1}^p \frac{f_i(x_i)}{\phi\big(\Phi^{-1}(F_i(x_i))\big)},
\label{eq:backtrans}
\end{equation}
where $\phi(\cdot)$ is the standard normal pdf. This estimator is implemented the \CRANpkg{lg}-package as demonstrated in Section \ref{chap:density}.

One particular feature enjoyed by the jointly normally distributed vector $X$ is that, for any partitioning $X = \left(X^{(1)}, X^{(2)}\right)$, the conditional distribution of $X^{(1)}|X^{(2)} = x^{(2)}$ is also normal. In fact, if $X \sim \mathcal{N}(\mu,\Sigma)$, and $\mu$ and $\Sigma$ is partitioned according to $\left(X^{(1)}, X^{(2)}\right)$ as
$$\mu = \begin{pmatrix} \mu_1 \\\mu_2 \end{pmatrix} \,\, \textrm{ and } \,\, \Sigma = \begin{pmatrix} \Sigma_{11} & \Sigma_{12} \\ \Sigma_{21} & \Sigma_{22} \end{pmatrix},$$
then $X^{(1)}|X^{(2)} = x^{(2)} \sim \mathcal{N}(\mu^*, \Sigma^*)$, where

\begin{align}
\mu^* &= \mu_1 + \Sigma_{12}\Sigma_{22}^{-1}\left(x^{(2)} - \mu_2\right) \label{eq:cond-mu} \\
\Sigma^* &= \Sigma_{11} - \Sigma_{12}\Sigma_{22}^{-1}\Sigma_{21}, \label{eq:cond-sig}
\end{align}
see e.g. \citet[chapter 4]{john:wich:2007}. \citet{otne:tjos:2018} demonstrate that this property may be translated into a corresponding local argument without modification. That is, if the joint density $f_X(\cdot)$ can be written on a locally Gaussian form

$$f_X(x) = \psi(x, \mu(x), \Sigma(x)),$$
then the conditional density of $X^{(1)}|X^{(2)} = x^{(2)}$ can also be written on the same locally Gaussian form with local parameters given by equations (\ref{eq:cond-mu}) and (\ref{eq:cond-sig}), except that all quantities are $x$-dependent. If we use the transformation technique described above together with simplification (\ref{eq:simp}), the local versions of equations (\ref{eq:cond-mu}) and (\ref{eq:cond-sig}) simplify to

\begin{align}
\mu^*(z) &= R_{12}(z)R_{22}(z)^{-1}z^{(2)},  \label{eq:cond-mu-trans}\\
\Sigma^*(z) &= R_{11}(z) - R_{12}(z)R_{22}(z)^{-1}R_{21}(z), \label{eq:cond-sig-trans}
\end{align}
where we, again, switch to $z$-notation in order to make it clear that these quantities are estimated on the standard normal $z$-scale. An estimator $\widehat f_{X^{(1)}|X^{(2)}}(\cdot|\cdot)$ of the conditional density $f_{X^{(1)}|X^{(2)}}(\cdot|\cdot)$ follows immediately from an expression corresponding to (\ref{eq:backtrans}), and the \CRANpkg{lg}-package provides functions for implementing this estimator in R. We describe the implementation of this functionality in Section \ref{chap:density}.

>>>>

\citet{tjos:huft:2013} propose the \emph{local Gaussian correlation} (LGC) as a new measure of statistical dependence between two stochastic variables $X_1$ and $X_2$, which has the following important property yet unrivaled in the literature: It can separate between positive and negative nonlinear dependence while still reducing to the ordinary Pearson correlation coefficient if $X_1$ and $X_2$ are jointly normally distributed. The R-package \CRANpkg{localgauss} \citep{bere:klep:tjos:2014} provides two important functions in this context; one that calculates the sample LGC based on observed values of $\left(X_1,X_2\right)$, and one that uses the estimated LGC to perform a local test of independence between $X_1$ and $X_2$ as described in detail by \citet{bere:tjos:2014}.

@@

Consider a random vector $X$ having the unknown probability density function $f_X\left(x\right)$. It is a standard task to estimate $f_X$ based on a random sample $X_1,\ldots, X_n$, and the statistical literature provides an abundance of methods to accomplish this. One may, for example, make the assumption that the unknown density function has a particular parametric form, $f_X \in F_{\theta}$, where $F_{\theta} = \left\{f(x;\theta), \theta \in \Theta\right\}$ is a family of probability density functions indexed by some parameter $\theta$, and where $\Theta$ is the parameter space. Under this assumption we will typically produce an estimate of the parameter $\theta$, written $\widehat \theta$, using the maximum likelihood method. The estimated probability density function is then given as $\widehat f_X\left(x\right) = f\left(x; \widehat\theta\right)$.

A different approach is to estimate $f_X\left(\cdot\right)$ without any prior parametric assumptions. The classical method for non-parametric density estimation is the kernel estimator 
$$\widehat f_X\left(x\right) = \frac{1}{nb}\sum_{i=1}^n K\left(\frac{X_i - x}{b}\right),$$
where $K$ is a symmetric density function (the kernel) and $b$ is a tuning parameter (the bandwidth) that controls the smoothness of the estimate $\widehat f_X\left(\cdot\right)$. See \citet{silv:1986} for an introduction to this topic. There is also a massive statistical literature on density estimation containing extensions and improvements to the classical methods to be used in various practical situations.

\citet{hjor:jone:1996} provide one such idea. They consider a parametric family $F_{\theta}$, but instead of searching for a single parameter value $\theta_0$, for which $f_X\left(x\right) = f\left(x;\theta_0\right)$ (or approximately so), they rather assert that different members of $F_{\theta}$ may approximate $f_X$ locally in different parts of its domain. In other words, they seek to estimate a parameter \emph{function} $\theta_0\left(x\right)$ for which $f_X\left(x\right) = f\left(x;\theta_0\left(x\right)\right)$ (or approximately so), and do this by maximizing a \emph{local likelihood function} in each point $x$:
\begin{align}
\widehat\theta\left(x\right) 
&= \arg\max_{\theta \in \Theta} L_n\left(\theta, x\right) \nonumber\\ 
&= \arg\max_{\theta \in \Theta} \frac{1}{nb}\sum_{i=1}^n K\left(\frac{X_i - x}{b}\right)\log f\left(X_i; \theta\right) - \int \frac{1}{b}K\left(\frac{y - x}{b}\right)f\left(y;\theta\right)\,\textrm{d}y,
\label{eq:loclik}
\end{align}
where, again, $K$ is a symmetric density function and $b$ is a bandwidth parameter that controls the smoothness of the estimate. The second term in the local likelihood function is a penalty that ensures that the estimated density $\widehat f_X\left(x\right) = f\left(x; \widehat\theta\left(x\right)\right)$ converges correctly to the true density function $f_X\left(x\right)$ as the sample size $n$ increases to infinity, and the bandwidth $b$ decreases towards zero. See \citet{hjor:jone:1996} for a detailed discussion about this construction.

\citet{tjos:huft:2013} consider the bivariate case $X = \left(X_1, X_2\right)$ and take $F_{\theta}$ to be the family of bivariate normal distributions consisting of densities on the form

\begin{align}
f\left(x;\theta\right) 
&= \psi\left(x_1, x_2;\mu_1, \mu_2, \sigma_1, \sigma_2, \rho\right)  \nonumber\\ 
&= \frac{1}{2\pi \sigma_1 \sigma_2\sqrt{1-\rho^2}} \nonumber \\
& \qquad \times\exp \left\{-\frac{1}{2(1-\rho^2)}\left(\frac{\left(x_1-\mu_1\right)^2}{\sigma_1^2}-2\rho\frac{\left(x_1-\mu_1\right)\left(x_2-\mu_2\right)}{\sigma_1 \sigma_2}+\frac{\left(x_2-\mu_2\right)^2}{\sigma_2^2}\right)\right\},
\label{eq:gaussian}
\end{align}
where $\theta = \left(\mu_1,\mu_2,\sigma_1,\sigma_2,\rho\right)$ is the vector of parameters. Using a sample $\left\{X_{1i}, X_{2i}\right\}$, $i = 1, \ldots, n$, they estimate $\theta$ locally in the point $x$ by maximizing the local likelihood function \eqref{eq:loclik}, producing
$$\widehat\theta\left(x\right) = \left(\widehat\mu_1\left(x\right), \widehat\mu_2\left(x\right), \widehat\sigma_1\left(x\right), \widehat\sigma_2\left(x\right), \widehat\rho\left(x\right)\right),$$
and take special interest in the estimated correlation function $\widehat\rho\left(x\right)$ (i.e. the LGC) because it serves as an attractive local measure of statistical dependence between $X_1$ and $X_2$. They show that the LGC reveals many types of nonlinear statistical dependence that are not captured by the ordinary (global) Pearson correlation coefficient. Furthermore, the LGC distinguishes between positive and negative dependence and reduces to the Pearson $\rho$ if $X_1$ and $X_2$ are jointly normal. We refer to \cite{tjos:huft:2013} for a detailed treatment of the theoretical foundations of the LGC as well as several examples, and rather present two simple illustrations at this point in order to demonstrate the concept.

\begin{figure}[t]
    \centering
    \begin{minipage}[b]{0.48\textwidth}
      \includegraphics[width=\textwidth]{gaussian-example}
      \subcaption{Estimated local Gaussian correlations between  \newline two jointly normal variables having correlation  \newline  equal to 0.5, based on 1000 observations.}
    \end{minipage}
    \begin{minipage}[b]{0.48\textwidth}
      \includegraphics[width=\textwidth]{returns-example}
      \subcaption{Local Gaussian correlation between daily returns from the CAC40 and FTSE100 stock indices, based on 1000 consecutive trading days.}
    \end{minipage}
    \caption{Two dependence maps}
    \label{fig:initial-example}
  \end{figure}

@@

Taking the LGC as a measure of dependence opens up for a number of possibilities to construct statistical tests. \citet{bere:tjos:2014} show that $\rho\left(x\right) \equiv 0$ implies that $X_1$ and $X_2$ are independent. They show further that independence between $X_1$ and $X_2$ implies $\rho\left(x\right) \equiv 0$ if the population values of the local mean and standard deviation functions satisfy the following conditions: $\mu_i\left(x_1, x_2\right) = \mu_i\left(x_i\right)$ and $\sigma_i\left(x_1, x_2\right) = \sigma_i\left(x_i\right)$ for $i=1,2$. Equivalence between independence and $\rho\left(x\right) \equiv 0$ holds in general if the observations have been suitably transformed according to a procedure presented later in this section. It follows then that departures from $\widehat\rho\left(x\right) \equiv 0$ may be taken as evidence against the hypothesis that $X_1$ and $X_2$ are statistically independent. \citet{bere:tjos:2014} formalize this notion by testing whether $\rho\left(x\right) \equiv 0$ for all $x\in S \subset \mathbb{R}^2$ using the test statistic

\begin{equation}
T_{n,b} = \int_S h\left(\widehat \rho\left(x\right) \right) \,\textrm{d}F_n(x)
\label{eq:test-statistic1}
\end{equation}
for some non-negative function $h$, for example $h\left(x\right) = x^2$ or $h\left(x\right) = |x|$. Critical values may be obtained by permutations of the data under the null hypothesis, and we demonstrate the implementation of this test using the \CRANpkg{lg}-package in Section \ref{chap:indtests}.

Consider next the stationary time series $\left\{X_t\right\}$. The autocorrelation function (ACF) $\rho_k = \rho\left(X_t, X_{t-k}\right)$ is a well known concept for describing the serial dependence in the time series, but the ACF is, again, only capable to completely capture \emph{linear} serial dependence. \citet{laca:tjos:2017} seek to remedy this by rather calculating the the local correlation between $X_t$ and $X_{t-k}$. This leads to a test for serial independence in the natural way. In fact, this work is mainly a theoretical exercise in order to accommodate time series dependence. Testing for independence between $X_t$ and $X_{t-k}$ using observations $\left\{X_t, X_{t-k}\right\}_{t = k+1}^T$ leads to the same test statistic \eqref{eq:test-statistic1} and bootstrap procedure as the test for independence between $X_1$ and $X_2$ that we described above.

\citet{laca:tjos:2018} extend this problem to test for serial cross-dependence between two time series $\left\{X_t, Y_t\right\}$ by measuring the LGC between $X_t$ and $Y_{t-k}$. Departures from $\widehat\rho\left(x,y\right) \equiv 0$ is again taken as evidence against independence, and the test statistic (\ref{eq:test-statistic1}) provides an aggregate measure of this discrepancy in the specified region $S$. In this case, however, we can not obtain replicates of the test statistic under the null hypothesis by simple permutations of the data. \citet{laca:tjos:2018} suggest rather two block bootstrap procedures to this end, using fixed and random block sizes respectively. The tests for serial dependence, and serial cross-dependence are both implemented in the \CRANpkg{lg}-package, as we demonstrate in Section \ref{chap:indtests}.

We find another application of the local Gaussian approximation in the work by \citet{stov:tjos:huft:2014}, who measure and test for financial contagion. They define contagion by "a significant increase in cross-market linkages after a shock to one country" \citep[p. 2223]{forb:rigo:2002} and employ the LGC to quantify this potential linkage. The authors estimate the LGC on a grid $\left\{x_1, x_2\right\}_k$, $k=1,\ldots,K$ along the diagonal $D=\left\{\left(x_1,x_2\right): x_1 = x_2\right\}$ \emph{before} and \emph{after} some critical event in the financial markets, denoted as the crisis (C) and the non-crisis (NC) periods respectively. They compare the two estimates using the following test statistic; 
$$T_{n,b}^D = \sum_{k=1}^K \left\{\widehat\rho_C\left(x_k, x_k\right) - \widehat\rho_{NC}\left(x_k, x_k\right) \right\}w\left(x_k,x_k\right),$$

where $w\left(\cdot, \cdot\right)$ is a weight function that serve the same purpose as the integration area $S$ in (\ref{eq:test-statistic1}). In this case, \citet{stov:tjos:huft:2014} show that a standard bootstrap will suffice in order to produce approximate replicates of $T_{n,b}^D$ under the null hypothesis of no financial contagion, and we demonstrate the implementation of this test using the \CRANpkg{lg}-package in Section \ref{chap:cont}.

Although the original work by \citet{hjor:jone:1996} provide a general framework for local likelihood density estimation using any $p$-variate parametric family as the local family, it is evident that the method may struggle in multivariate applications much in the same way as the kernel density estimator does. This is a consequence of the curse of dimensionality, the effect of which is sought remedied by an algorithm provided by \citet{otne:tjos:2017}. The idea is to fit the $p$-variate normal distribution $\psi\left(\mu, \Sigma\right)$ locally, where $\mu$ is the vector of $p$ expectations, and $\Sigma$ is the $p\times p$ covariance matrix (to which the correlation matrix $R$ corresponds), but under the following structural simplifications:

\begin{align}
\mu_i\left(x\right) &= \mu_i\left(x_1, \ldots, x_p\right) \stackrel{\textrm{def}}{=} \mu_i\left(x_i\right) \label{eq:simp-mu}\\
\sigma_i\left(x\right) &= \sigma_i\left(x_1, \ldots, x_p\right)  \stackrel{\textrm{def}}{=} \sigma_i\left(x_i\right) \label{eq:simp-sig} \\
\rho_{ij}\left(x\right) &= \rho_{ij}\left(x_1, \ldots, x_p\right) \stackrel{\textrm{def}}{=} \rho_{ij}\left(x_i, x_j\right). \label{eq:simp-rho}
\end{align}
\citet{otne:tjos:2017} estimate the local parameters above by first obtaining univariate marginal locally Gaussian fits (eqs. \ref{eq:simp-mu} and \ref{eq:simp-sig}), and then pairwise bivariate locally Gaussian fits (eq. \ref{eq:simp-rho}). In the second step the estimates $\widehat\mu_i\left(x_i\right)$, $\widehat\mu_j\left(x_j\right)$, $\widehat\sigma_i\left(x_i\right)$ and $\widehat\sigma_j\left(x_j\right)$ are kept fixed in the estimation of the pairwise local correlation. They argue further that the following transformation technique will produce better density estimates in many situations. The motivation for introducing the simplifications defined in equations \ref{eq:simp-mu}-\ref{eq:simp-rho} can be compared to the practical advantages of estimating additive regression models, where $\textrm{E}\left(Y\right) = f\left(x_1, \ldots, x_p\right) \stackrel{\textrm{def}}{=} f_1\left(x_1\right) + \cdots + f_p\left(x_p\right)$.

Denote by $F_i\left(x_i\right)$, $i = 1,\ldots, p$ the marginal distribution functions of the stochastic vector $X$, and by $\widehat F_i\left(x_i\right) = n^{-1}\sum_{i=1}^n 1\left(X_i \leq x_i\right)$ their empirical counterparts. They then estimate the density $f_Z\left(z\right)$ of the vector $Z = \left\{\Phi^{-1}\left(F_i\left(X_i\right)\right)\right\}_{i=1,\ldots,p}$, in practice approximated by

\begin{equation}
\widehat Z = \left\{\Phi^{-1}\left(\widehat F_i\left(X_i\right)\right)\right\}_{i=1,\ldots,p}, 
\label{eq:trans}
\end{equation}
and where $\Psi\left(\cdot\right)$ is the univariate standard normal cdf. In that case, they simplify the estimation problem even further and fix

\begin{equation}
\mu_i\left(z_i\right) \stackrel{\textrm{def}}{=} 0 \,\, \textrm{ and } \sigma_i\left(z_i\right) \stackrel{\textrm{def}}{=} 1, \,\, i = 1,\ldots,p,
\label{eq:simp}
\end{equation}
so that the only parameter functions left to estimate are the pairwise local Gaussian correlations $R\left(z\right) = \left\{\rho_{ij}\left(z_i, z_j\right)\right\}_{i<j}$. We use the notation $Z$, $z_i$ and $z_j$ to signify that the estimation is performed on the (approximate) standard normal scale, or $z$-scale for short. We can then estimate the joint density $f_Z\left(z\right)$ of $Z$ as
\begin{equation}
\widehat f_Z\left(z\right) = \psi\left(z; \mu\left(z\right) = 0, \sigma\left(z\right) = 1, R = \widehat R\left(z\right)\right),
\label{eq:transformed-density}
\end{equation}
where $\mu\left(z\right) = \left\{\mu_i\left(z\right)\right\}$ and $\sigma\left(z\right) = \left\{\sigma_i\left(z\right)\right\}$ for $i=1,\ldots,p$, and then substitute $f_Z$ for $\widehat f_Z$ in the following relation obtained by \citet{otne:tjos:2017} in order to estimate the unknown density $f_X$:
\begin{equation}
f\left(x\right) = f_{Z}\big(\Phi^{-1}\left(F_1\left(x_1\right)\right), \ldots, \Phi^{-1}\left(F_p\left(x_p\right)\right)\big) \times \prod_{i=1}^p \frac{f_i\left(x_i\right)}{\phi\big(\Phi^{-1}\left(F_i\left(x_i\right)\right)\big)},
\label{eq:backtrans}
\end{equation}
where $\phi\left(\cdot\right)$ is the standard normal pdf. This estimator is implemented the \CRANpkg{lg}-package as demonstrated in Section \ref{chap:density}.

One particular feature enjoyed by the jointly normally distributed vector $X$ is that, for any partitioning $X = \left(X^{\left(1\right)}, X^{\left(2\right)}\right)$, the conditional distribution of $X^{\left(1\right)}|X^{\left(2\right)} = x^{\left(2\right)}$ is also normal. In fact, if $X \sim \mathcal{N}\left(\mu,\Sigma\right)$, and $\mu$ and $\Sigma$ is partitioned according to $\left(X^{\left(1\right)}, X^{\left(2\right)}\right)$ as
$$\mu = \begin{pmatrix} \mu_1 \\\mu_2 \end{pmatrix} \,\, \textrm{ and } \,\, \Sigma = \begin{pmatrix} \Sigma_{11} & \Sigma_{12} \\ \Sigma_{21} & \Sigma_{22} \end{pmatrix},$$
then $X^{\left(1\right)}|X^{\left(2\right)} = x^{\left(2\right)} \sim \mathcal{N}\left(\mu^*, \Sigma^*\right)$, where

\begin{align}
\mu^* &= \mu_1 + \Sigma_{12}\Sigma_{22}^{-1}\left(x^{\left(2\right)} - \mu_2\right) \label{eq:cond-mu} \\
\Sigma^* &= \Sigma_{11} - \Sigma_{12}\Sigma_{22}^{-1}\Sigma_{21}, \label{eq:cond-sig}
\end{align}
see e.g. \citet[chapter 4]{john:wich:2007}. \citet{otne:tjos:2018} demonstrate that this property may be translated into a corresponding local argument without modification. That is, if the joint density $f_X\left(\cdot\right)$ can be written on a locally Gaussian form 
$$f_X\left(x\right) = \psi\left(x, \mu\left(x\right), \Sigma\left(x\right)\right),$$
then the conditional density of $X^{\left(1\right)}|X^{\left(2\right)} = x^{\left(2\right)}$ can also be written on the same locally Gaussian form with local parameters given by equations (\ref{eq:cond-mu}) and (\ref{eq:cond-sig}), except that all quantities are $x$-dependent. If we use the transformation technique described above together with simplification (\ref{eq:simp}), the local versions of equations (\ref{eq:cond-mu}) and (\ref{eq:cond-sig}) simplify to
